$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-08 paper/paper.tex
Update the explanations on the typing derivation example

We present an example that highlights the ways in which graded types can be used to prevent mistakes in the definition of \picalc{} processes.
The syntax uses Bruijn indices as variable references and is introduced in detail in \autoref{syntax}.
In short: $\new{} P$ creates a new channel and makes it available at index $0$ in $P$; $\comp{P}{Q}$ composes $P$ and $Q$ in parallel; $\recv{x}{}P$ receives along channel $x$ and makes the data available at index $0$ in $P$; and $\send{x}{y}{P}$ sends $y$ over channel $x$ and continues with $P$.
Both variable references and processes are indexed by the number of free variables they can refer to (for any natural number $n$, $\Var_n$ and $\Process_n$, respectively).
For the sake of clarity, we define the example subprocesses separately and parametrise them by the channels on which they operate.

@@

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks for linear and graded types, the processes below have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{sync}$ consumes one input multiplicity on the channels shared with $\func{send}$, and two output multiplicities on the channel shared with $\func{recv}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation}.

@@

\begin{flalign*}
  & \func{\vdash send} \; \types{\gamma \comma \channel{\channel{\unit}{\func{\omega}}}{\lz}}{\Gamma \comma (l \comma \suc r)}{\func{send} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash send} \; = \; \new{} \; \_ \; \_ \; \zero \; \send{(\suc \constr{0})}{\constr{0}} \PO \\
  & \\
  & \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (\suc \suc l \comma r)}{\func{recv} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
\end{flalign*}

We expect both input channels to be graded and capable of performing input once.
The output channel must be capable of performing output twice.
Notice that the well-typedness proof has no choice but to blindly follow the syntax.

@@

Finally, we put the four processes together in parallel.
It is here that the well-typedness proof must make a choice as to what types and usage annotations to instantiate new channels with.
In this case, these types and usage annotations were determined by the types of the previous subproofs, and we can thus omit most of them with $\_$ and have Agda find them.
We use $\func{\omega}$ as a shortcut for $\one \comma \one$ in an algebra of shared types.

>>>>

We present an example that highlights the ways in which graded types can be used to prevent mistakes in the definition of \picalc{} processes.
The syntax uses de Bruijn indices for variables and is introduced in detail in \autoref{syntax}.
In short: $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation $P$; $\comp{P}{Q}$ composes $P$ and $Q$ in parallel; $\recv{x}{}P$ receives along channel $x$ and makes the data available at index $0$ in the continuation $P$; and $\send{x}{y}{P}$ sends $y$ over channel $x$ and continues with $P$.
Both variable references and processes are indexed by the number of free variables they can refer to (for any natural number $n$, $\Var_n$ and $\Process_n$, respectively).
For the sake of clarity, we define the example subprocesses separately and parametrise them by the channels on which they operate.

@@

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous three processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks to linear and graded types, the processes below can have their interactions safely restricted: both $\func{send}$ processes consume $0$ input and $1$ output multiplicities; $\func{recv}$ consumes $2$ input and $0$ output multiplicities.
Dually, the process $\func{sync}$ consumes one input multiplicity on the channels shared with $\func{send}$, and two output multiplicities on the channel shared with $\func{recv}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation}.

@@

The $\func{recv}$ process receives data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$.
After receiving twice, the process ends, so we should not be left with unused multiplicities: $u$ needs to be $\lz$.
We will use graded types to keep track of the exact number of times communication happens.
Whatever the input multiplicity of the channel, we will consume $2$ of it and leave the remaining as leftovers.
Once the types are given, the typing derivation is completely syntax directed.

The $\func{send}$ process sends data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$.
We instantiate the type of transmitted data $t$ to $\channel{\unit}{\func{\omega}}$, as that is something we can construct.
As by the type of the process $\func{recv}$, the transmitted multiplicities $u$ are $\lz$.
We will transmit once, thus using a single output multiplicity, and leave the rest as leftovers.

\begin{flalign*}
  & \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (\suc \suc l \comma r)}{\func{recv} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
  & \\
  & \func{\vdash send} \; \types{\gamma \comma \channel{\channel{\unit}{\func{\omega}}}{\lz}}{\Gamma \comma (l \comma \suc r)}{\func{send} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash send} \; = \; \new{} \; \_ \; \_ \; \zero \; \send{(\suc \constr{0})}{\constr{0}} \PO \\
\end{flalign*}

Dually, the process doing the synchronisation expects input multiplicities for the channels shared with $\func{send}$ and output multiplicities for the channel shared with $\func{recv}$.
The typing derivation here is uniquely determined by the type and syntax of the process.

@@

We can now put these processes together and omit specifying the types of the channels that they share: Agda can infer them from the types of the processes.
